---
title: "My Project"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Video Game Sales Analysis

Mason Gliege, Ewan Farnum, Ishan Nichols

## Introduction

The video game industry is a multi-billion dollar global market. Our group aims to study the historical trends of how video games perform in order to identify relationships between factors like critic score, genre, total sales, and more.

Here are some questions we aim to answer, including but not limited to:

-   What are the most popular genres of video games?

-   How are video game sales distributed across regions of the global market?

-   Do some consoles have higher average video game sales than others?

-   What video game publishers are most dominant? How has this changed?

-   Is there a correlation between a game's sales and its critic score? What other factors might influence this relationship?

## Data and Variables

Our project utilized the "Video Game Sales & Industry Data (1980 - 2024)" dataset from Kaggle. This 8.3 MB dataset contains tens of thousands of video games, where each row is a game, and contains information such as its total sales, critic score, genre, etc. The complete list of variables can be found below.

**Variables:**

title: Game title

console: The game was released for the console

genre: Genre of the game

publisher: Publisher of the game

developer: Developer of the game

critic_score: Metacritic score (out of 10)

total_sales: Global sales of copies in millions

na_sales: North American sales of copies in millions

jp_sales: Japanese sales of copies in millions

pal_sales: European & African sales of copies in millions

other_sales: Rest of world sales of copies in millions

release_date: Date the game was released on

last_update: Date the data was last updated

## Cleaning

The Kaggle dataset was very easy to work with, but still required some basic cleaning steps before we could work with it.

First, we read the dataset and stored it as a dataframe.

The data had some null sales numbers, so we replaced them with zeros.

There were also null values for critic scores. Since there is no way to estimate scores on games using the existing data, we removed all rows with no score. Additionally, there were null values in total sales so those were also removed

```{r}
library(tidyverse)
df <- read.csv("video_games_raw.csv")

str(df)

df_clean <- df %>%
 mutate(
    jp_sales = replace_na(jp_sales, 0),
    na_sales = replace_na(na_sales, 0),
    pal_sales = replace_na(pal_sales, 0),
    other_sales = replace_na(other_sales, 0)
  ) %>%
  drop_na(critic_score, total_sales)

```

## Next, we split the games into different categories by sales.

Small Games: under 100k sales

Medium Games: between 100k and 1 million sales

Large Games: between 1 million and 10 million sales

Very_Large_Games: over 10 million sales

```{r}
# Filter for games with less than 100k sales
small_games <- df_clean %>% 
  filter(total_sales < 0.1)

# Filter for games between 100k and 1 million sales
medium_games <- df_clean %>% 
  filter(total_sales >= 0.1 & total_sales < 1)

# Filter for games between 1 million and 10 million sales
large_games <- df_clean %>% 
  filter(total_sales >= 1 & total_sales < 10)

# Filter for games with 10 million or more sales
very_large_games <- df_clean %>% 
  filter(total_sales >= 10)


```

## Top 25 Best Selling Games

```{r}
top_games <- df_clean %>%
  group_by(title)%>%
  summarise(total_sales = sum(total_sales))%>%
  arrange(desc(total_sales)) %>%
  head(25)

ggplot(top_games, aes(x = reorder(title, total_sales), y=total_sales)) +
  geom_col(fill = "steelblue") +
  coord_flip()+
  labs(
    title = "Top 25 Best-Selling Games",
    x = "Game",
    y = "Total Sales (millions)",
    fill = "Console"
  )
```

From the chart we can see that GTA V is the best selling game by a large margin with almost 50 million sales.
The top games are dominated by Call of Duty and GTA franchises, making up 13 of the top 25 games. Additionally, FIFA makes up another 4 games, so 3 franchises make up 17 out of the top 25 games. From this we can see that franchise recognition is a major component of success in games.

## Critic Score vs Sales

```{r}
ggplot(df_clean, aes(x = critic_score, y = total_sales)) +
  geom_point(alpha =0.3)+
   labs(
    title = "Critic Score vs. Total Sales",
    x = "Critic Score",
    y = "Total Sales (millions)"
  ) 

cor(df$critic_score, df$total_sales, use = "complete.obs")
```

From the correlation coefficient of 0.28, we can see that critic score and total sales have a weak positive correlation. The graph shows that as the critic score increases, the total sales in millions increases slightly, but there is not a very obvious direct relationship. What we do see, however, is that all games with over 10 million sales have a critic score of at least 7. We will revisit the correlation later.

## Sales by Genre

```{r}
genre_summary <-df_clean %>%
  group_by(genre) %>%
  summarise(
    avg_sales = mean(total_sales),
    count = n()
  ) %>%
  arrange(desc(avg_sales))

genre_summary

ggplot(genre_summary, aes(x = reorder(genre, avg_sales), y = avg_sales))+
  geom_col(fill = "steelblue")+
  coord_flip()+
  labs(
    title = "Average Sales by Genre",
    x = "Genre",
    y = "Average Total Sales (millions)"
  ) 
```

From the bar graph, we can see which genre of video games sell more on average. The bar graph shows this to be clearly sandbox, but this is not a reliable conclusion as there is only one sandbox game included in the dataset. The true most popular games which have a statistically significant sample size are shooter, action-adventure, and sports. The least popular games are strategy, puzzle, and platform. Again the graph shows genres like Visual Novel, MMO, and Board Game, but their sample sizes are too low to draw any conclusions.

## Average Sales by Console

```{r}
console_summary <- df_clean %>%
  group_by(console) %>%
  summarise(
    avg_sales = mean(total_sales),
    count = n()
  )%>%
  filter(count >= 20) %>%
  arrange(desc(avg_sales))

ggplot(console_summary, aes(x = reorder(console, avg_sales), y = avg_sales)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
   labs(
    title = "Consoles by Average Total Sales",
    x = "Console",
    y = "Average Total Sales (millions)"
  ) 
```

From the bar graph, we can see that the consoles with the highest average total game sales are the PS4 in first by far, with about 1 million average sales higher than the Xbox One, and then the PS3, PS2, Xbox 360 are next and all relatively close to each other. From the chart we can see that Playstation and Xbox lead average sales by a large margin compared to other consoles.

## Sales by Region

```{r}
regional_sales <- df_clean %>%
  summarise(
    `North America` = sum(na_sales),
    `Japan` = sum(jp_sales),
    `Europe/Australia` = sum(pal_sales),
    `Other` = sum(other_sales)
  )%>%
  pivot_longer(everything(), names_to = "region", values_to= "sales")

ggplot(regional_sales, aes(x= reorder(region, sales), y = sales, fill = region))+
  geom_col()+
  coord_flip()+
   labs(
    title = "Total Sales by Region",
    x = "Region",
    y = "Total Sales (millions)"
  )+
  theme(legend.position = "none")
```

From the chart we can see that North America has the largest market for video games with around 1.5 billion total sales, next is Europe and Australia with around 1 billion.

## Sales by game size category

```{r}
small_games$sales_category <- "Small (<100k)"
medium_games$sales_category <- "Medium (100k-1M)"
large_games$sales_category <- "Large (1M-10M)"
very_large_games$sales_category <- "Very Large (10M+)"

all_games <- bind_rows(small_games, medium_games, large_games, very_large_games)

all_games$sales_category <-factor(all_games$sales_category,
                                  levels = c("Small (<100k)", "Medium (100k-1M)", "Large (1M-10M)", "Very Large (10M+)"))

ggplot(all_games, aes(x = sales_category, fill = sales_category))+
  geom_bar()+
  labs(
    title = "Distribution of Games by Sales Category",
    x = "Sales Category",
    y = "Number of Games"
  )
  
```

The vast majority of games fit into the medium category with 100k to 1M sales. The medium category has more games than any of the other categories combined.
Very large games that sell over 10 million copies make up a very small portion of the total games.
From this we can see that a small fraction of games generate extreme sales while the majority of games sell moderately well.

## Score by game size

```{r}
ggplot(all_games, aes(x = sales_category, y = critic_score, fill = sales_category)) +
  geom_boxplot() +
  labs(
    title = "Critic Score Distribution by Sales Category",
    x = "Sales Category",
    y = "Critic Score"
  ) 
```

The median critic score increases slightly as game size increases. Very large games have the smallest spread, indicating that blockbuster games are almost always highly reviewed. Small games have the largest spread and lowest median, displaying that low selling games tend correspond to lower reviews.

## Skepticism of Correlation: Critic Score vs Total Sales (Millions)

Earlier, we found a relatively weak, positive correlation of **0.28** between Critic Score and Total Sales. What if this correlation is weak because it takes into account **all** video games, which have a variety of factors influencing sales? For example, certain genres may have fans that are more critical of the game than others. A racing game might be criticized by how well the physics engine runs, while an acion game might be praised by its cinematics.

Since different genres have different criteria for critic score, we hypothesized that narrowing the correlation to a few specific genres would lead to a stronger correlation between the critic score and total sales.

## Investigating correlation of Critic Score vs Total Sales in the most popular genres

In this code, we identified the top 5 most popular genres by the number of games in those genres. Then, made a faceted scatterplot showing the trendline and r-value of the Critic Score vs Total Sales in each genre.

```{r}
#identify the top 5 genres of games
topGenres <- df_clean |> count(genre) |> top_n(5,n) |> pull(genre)

#filter with top genres
q1 <- df_clean |> filter(genre %in% topGenres)

#calculate correlations by genre
correlations <- q1 |> group_by(genre) |> summarise(
  corVal = cor(critic_score, total_sales, use = "complete.obs"),
  .groups = 'drop'
) |> mutate(
  label = paste0("r = ", round(corVal,2))
)


#plot with facet wrap and correlation
q1 |> ggplot(aes(x = critic_score, y = total_sales, color = genre)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = F, color = "black") + #trend line gets added
  geom_text(data = correlations, #this adds the corVal in the rigth spot for each genre
            aes(x = -Inf, y = Inf, label = label), 
            hjust = -0.2, vjust = 1.5, 
            color = "black", size = 5, fontface = "bold",
            inherit.aes = FALSE) +
  facet_wrap(~genre) +
  theme_minimal() +
  labs(
    title = "Critic score vs total sales across top genres",
    x = "Critic Score",
    y = "Total Sales (millions)"
  )


```

This revealed that the top 5 video game genres all have correlations that, while still relatively weak, are higher than the correlation from all games combined (up to 0.37). We believe that this is due to specific, well known genres having predictable patterns, while smaller games in lesser known genres create a lot of "noise" that throws off the r-value.

## Sales Over Time

```{r}
df_clean <- df_clean %>%
  mutate(year = as.integer(str_sub(release_date, -4)))%>%
  filter(!is.na(year), year >= 1980, year <= 2023)

sales_by_year <- df_clean %>%
  group_by(year) %>%
  summarise(total_sales = sum(total_sales))%>%
  filter(year <= 2011)

ggplot(sales_by_year, aes(x = year, y = total_sales)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    title = "Total Game Sales Over Time",
    x = "Year",
    y = "Total Sales (millions)"
  ) 
```

Looking at the chart total game sales rose rapidly through the early 2000s peaking around 2011 which reflects the height of the PS3 and Xbox 360 era and dominance of franchises like Call of Duty. Years after 2019 are excluded due to incomplete data coverage that does not reflect the actual total game sales of those years. The overall trend shows a clear growth through the 2000s and into the 2010s. 

## Top 10 Publishers by Average Sales

```{r}
publisher_summary <- df_clean %>%
  group_by(publisher) %>%
  summarise(
    avg_sales = mean(total_sales),
    count = n()
  ) %>%
  filter(count >= 10) %>%
  arrange(desc(avg_sales)) %>%
  head(10)

ggplot(publisher_summary, aes(x = reorder(publisher, avg_sales), y = avg_sales)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Publishers by Average Sales",
    x = "Publisher",
    y = "Average Sales per Game (millions)"
  )
```

We can see that Rockstar Games is by far the best selling publisher with over double the average sales of Microsoft in second. This reflects with the imense success of GTA along with their other games. We can see the next 9 publishers all have comparable average sales.

## Regional Sales by Genre

```{r}
regional_genre <- df_clean %>%
  group_by(genre) %>%
  summarise(
    `North America` = sum(na_sales),
    `Japan` = sum(jp_sales),
    `Europe/Australia` = sum(pal_sales),
    `Other` = sum(other_sales)
  ) %>%
  pivot_longer(-genre, names_to = "region", values_to = "sales")

ggplot(regional_genre, aes(x = reorder(genre, sales), y = sales, fill = region)) +
  geom_col(position = "fill") +
  coord_flip() +
  labs(
    title = "Regional Sales Breakdown by Genre",
    x = "Genre",
    y = "Proportion of Sales",
    fill = "Region"
  )
```

From the chart we can see that North America has the largest proportion of sales in almost all genres, which coorelates with the large total proportion North American sales make up of all games. Shooter and Action games are especially dominated by NA and European/Australian sales which reflects the popularity of franchises like Call of Duty in those regions. Japan has their largest proportion in party games which reflects the popularity Nintendo with games like Mario Kart.

## Critic Score Distribution by Genre

```{r}
valid_genres <- genre_summary %>% filter(count >= 10)

ggplot(df_clean %>% filter(genre %in% valid_genres$genre), aes(x = reorder(genre, critic_score, FUN = median), 
                     y = critic_score, fill = genre)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Critic Score Distribution by Genre",
    x = "Genre",
    y = "Critic Score"
  ) +
  theme(legend.position = "none")
```

Sports and Strategy games have the highest median scores, while action and platformer games have the lowest with large spreads, showing more variability in those genres. The genres that had the highest average sales, Shooter and Action-Adventure games, sit in the middle of the critic score rankings, showing that commercial success and critic score do not always correlate.

## The shift in dominance of top 5 publishers over time

One important aspect of studying trends in the video game industry is not only the game genres, but also the publishers who released them. We decided to analyze the top 5 publishers over a 20 year span from 2000-2019. The top 5 were chosen by which publishers had the highest sum of total sales over that 20 year span. The figure below shows how different publishers dominate the market by displaying their total sales for each year.

```{r}
#first identify which publishers are the top 5
topPublishers <- df_clean |> group_by(publisher) |> summarise(total = sum(total_sales)) |> top_n(5, total) |> pull(publisher)
#calculate the yearly sales of each publisher
q2 <- df_clean |> filter(publisher %in% topPublishers) |> filter(year >=2000 & year < 2019) |> group_by(year, publisher) |> summarise(yearly_sales = sum(total_sales), .groups = 'drop')

#plotting the data over time with geom line and point
q2 |> ggplot(aes(x = year, y = yearly_sales, color = publisher)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Change in publisher dominance over time",
    x = "Release year (2000-2019)",
    y = "Total sales (millions)",
    color = "Publisher"
  )
```

It is interesting how each of these publishers can have dramatically different sales from year to year, and it's hard to determine one publisher that dominates over this large 20 year span. In fact, many of the spikes in sales could be attributed to the release of major games. Some notable games that contribute to these spikes include:

-   Call of Duty: Modern Warfare 2 (Activision, 2009)

-   Grand Theft Auto V (Rockstar Games, 2013

-   Madden 12 + Battlefield 3 (Electronic Arts, 2011)

## Conclusions and Recommendations

-   Publishers should focus on the North American market, as the majority of sales come from there.

-   Studios should prioritize developing sandbox, shooter, action-adventure, and sports games, as those are the best-selling genres.

-   Critic scores should not be used as a complete and accurate measure of a game's success, the correlation is positive but weak.

-   Publishers should continue to constantly push for big new releases, as the dominant player in the market is constantly changing.

## Limitations

-   While our data was suitable for this project, the total sales for each game is only a single number. A dataset which contains the sales for each game during periods of time (months, quarters, years, etc.) would produce more detailed and accurate results of how sales change

-   The data is quite comprehensive of what video games released until about 2020. From 2020-2024, it seemed that there were many notable game releases that weren't accounted for.

-   Reliance on just this data source limited our ability to show how sales change. There are other external factors contributing to game sales that weren't present in the dataset. For example, an economic recession might lead to lower overall sales.

-   This dataset limited releases to consoles (for the most part), and didn't account for much activity from sites like steam

## Future Research

If we were to continue this project, there are a number of things we could expand on:

-   Combining this dataset with a dataset of global economic trends could prove that sales fluctuate depending on the condition of the regional or global economy

-   Obtaining data from Steam would be very useful since PC gaming is a very large market that's mostly excluded from this dataset. There's a higher amount of smaller "indie" games which likely have different sales patterns due to them being in a much smaller market

-   Machine learning models like logistic regression or random forest could be used to **predict** how well a game will perform (critic score, sales, etc) based on variables like its genre, console, etc.
